Regionally Additive Models: Explainable-by-design models minimizing feature interactions

### 2023-06-15 report/report.tex
completed the algorithmic part

For minimizing Eq.~\eqref{eq:SE}, we set up a CART-like algorithm, as proposed by \citep{herbinger2023decomposing}.
The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_identification}.
The basic idea is for each feature $s \in \{1, \ldots, D\}$, to exhaustively search level-by-level, for
all possible splits before choosing the one that minimizes the interaction.
For the first level, we search

\begin{algorithm}
\caption{DALE-based Region Detection}
\label{alg:subregion_identification}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\BlankLine
\Input{X, J, T}
\BlankLine
\Output{optimal\_splits}
\BlankLine
splits[s, t] $\gets$ None\;
positions[s, t] $\gets$ None\;
feat\_c[s, t] $\gets$ None\;
\For{$s = 1$ to $D$}{
    X\_list $\gets [X]$\;
    J\_list $\gets [J]$\;
    \For{$t = 1$ to $T$}{
            L, p, c, X\_list, J\_list $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
            splits[s, t] $\gets$ L\;
            position[s, t] $\gets$ p\;
            feat\_c[s, t] $\gets$ c\;
        }
    }
}
\end{algorithm}

@@

% Initialize variables
\BlankLine
positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
L $\gets$ []\; % \Comment{List to store interaction values}
\For{p \textnormal{in} positions}{
    X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, p, is\_cat)\;
    L $\gets$ GetInteraction(X\_split, J\_split)\;
}
split\_position = argmin(L)\;
position $\gets$ positions[split\_position]\;
X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, position, is\_cat)\;
\Return{L, position, X\_split, J\_split}
\BlankLine
\end{algorithm}

@@

    \subsubsection{Algorithmic Details of Subregion Detection}

>>>>

For minimizing Eq.~\eqref{eq:SE}, we set up a CART-like algorithm, as proposed by \citep{herbinger2023decomposing}.
The backbone of the algorithm is described in Algorithm~\ref{alg:subregion_detection};
for each feature $s \in \{1, \ldots, D\}$, we exhaustively search level-by-level, for the optimal split.
For example, suppose that the optimal first-level split for some feature $s$ is
$x_c \pm p_1$ where $x_c$ is a numerical feature.
The first-level split adds to the RAM to terms for feature $x_s$, $f_{s1}(x_s) \when{x_c < p_1}$ and $f_{s2}(x_s) \when{x_c \geq p_1}$,
and splits the dataset into two subsets, $X_1 = \{ \xb^i : x_s^i \leq x_c + p_1 \}$ and $X_2 = \{ \xb^i : x_s^i > x_c + p_1 \}$.
Afterwards, suppose that the optimal second-level split for feature $s$ is $x_c' = (\neq) p_2$ for some categorical feature $x_c'$.
The second-level split adds to the RAM four terms for feature $x_s$, $f_{s1}(x_s) \when{x_c < p_1 \land x_c' = p_2}$,
$f_{s2}(x_s) \when{x_c < p_1 \land x_c' \neq p_2}$, $f_{s3}(x_s) \when{x_c \geq p_1 \land x_c' = p_2}$ and $f_{s4}(x_s) \when{x_c \geq p_1 \land x_c' \neq p_2}$.

The output of algorithm~\ref{alg:subregion_detection} is a set of $T$ splits per feature $s$.
Since splitting does guarantee an important decrease in the loss function, we refine the splits by
pruning all splits that do not decrease the loss function by more than a threshold $\epsilon$.

\begin{algorithm}
\caption{Detection of Subregions using DALE}
\label{alg:subregion_detection}
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\BlankLine
\Input{Data matrix $X$, Jacobian matrix $J$, Maximum Depth $T$}
\BlankLine
\Output{Optimal splits}
\BlankLine
Initialize empty arrays for \texttt{loss}, \texttt{position}, and \texttt{feature\_c}
\BlankLine
\For{$s = 1$ to $D$}{
X\_list, J\_list = [X], [J] \tcp*{Init lists}
\For{$t = 1$ to $T$}{
    \tcc{Find best split}
    L, p, c, X\_list, J\_list $\gets$ BestSplit(X\_list, J\_list, s, is\_cat)\;
    \tcc{Store best split}
    loss[s, t], position[s, t], feature\_c[s, t] $\gets$ L, p, c
}
}
\Return{loss, position, feature\_c}
\end{algorithm}

@@

% Initialize variables
\BlankLine
L[c, p] $\gets$ None\; % \Comment{List to store interaction values}
\For{c $s = 1$ to $D$ \textnormal{if} c $\neq$ s}{
    is\_cat $\gets$ IsCategorical(X\_list, c)\;
    positions $\gets$ GetPositions(X\_list, c, is\_cat)\;
    L\_pos $\gets$ []\; % \Comment{List to store interaction values}
    \For{p \textnormal{in} positions}{
        X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c, p, is\_cat)\;
        L $\gets$ GetInteraction(X\_split, J\_split)\;
    }
}
L\_min $\gets$ min(L)\;
c\_min, p\_min = argmin(L)\;
is\_cat $\gets$ IsCategorical(X\_list, c\_min)\;
X\_split, J\_split $\gets$ SplitDataset(X\_list, J\_list, c\_min, p\_min, is\_cat)\;

\Return{L\_min, p\_min c\_min, X\_split, J\_split}
\BlankLine
\end{algorithm}

\paragraph{Computational Complexity}

Algorithm~\ref{alg:subregion_detection} has a computational complexity of $\mathcal{O}(D \cdot T)$.
and Algorithm~\ref{alg:single_feature_subregion} has a computational complexity of $\mathcal{O}(D-1 \cdot P \cdot N)$
where $D$ is the number of feature, $N$ is the number of samples, $T$ is the maximum depth of the tree and $P$ is the number of query postitions.
The computational complexity of the proposed method is therefore $\mathcal{O}(D \cdot (D-1) \cdot T \cdot P \cdot N)$.
However, in practice, the number of query positions $P$ and the maximum depth $T$ are small numbers.
Therefore, the computational complexity of the proposed method is $\mathcal{O}(D^2 \cdot N)$.

@@

\subsection{Algorithmic Details of Subregion Detection}
